---
title: "StatLearning 2024/25"
output: html_document
---

# Lab activity 4: Resampling Methods


##  4.1. Warming up: sampling with and without replacement  

Consider a simple dataset of numbers from 1 to 30:

```{r chunk1}
n=30
X=seq(from=1, to=n, by=1)
X
```

We begin by using the
`sample()` function to draw a sample containing a prortion $\alpha=60\%$ of units of $X$; see `?sample` for details. The option *replace* means with replacement (*replace=T*)
or without replacement (*replace=F*). By default, in the `sample()` function samples without replacement (*replace=F*). When we set  *replace=T*, some values appear more than once. 

```{r chunk2}
alpha=0.6
sample(X, alpha*n, replace=T) # sampling with replacement
sample(X, alpha*n, replace=T) 
sample(X, alpha*n, replace=T) 
sample(X, alpha*n, replace=F) # sampling without replacement
sample(X, alpha*n, replace=F)
sample(X, alpha*n, replace=F)
sample(X, alpha*n, replace=F)
```

We remark that each time we run the `sample()` function we get a different subset. 

We use the `set.seed()` function in order to set a  for `R`'s random number generator. It is generally a good idea to set a random seed so that the results obtained can be reproduced precisely at a later time.

```{r chunk3}
set.seed(1500)
sample(X, alpha*n, replace=TRUE) 
set.seed(1500)
sample(X, alpha*n, replace=TRUE) 
set.seed(1500)
sample(X, alpha*n, replace=F)
set.seed(1500)
sample(X, alpha*n, replace=F)
```

## 4.2. Selection of the training, validation and test sets

Consider the `Default`data in the `ISLR2` package concerning a a data frame with 10000 observations on the following 4 variables:

*default*: A factor with levels No and Yes indicating whether the customer defaulted on their debt </br>
*student*: A factor with levels No and Yes indicating whether the customer is a student </br>
*balance*: The average balance that the customer has remaining on their credit card after making their monthly payment </br>
*income*:  Income of customer </br>


```{r chunk4}
library(ISLR2)
attach(Default)
N<-nrow(Default)
str(Default)
```

Let's explore the dataset with respect to the target variable *default*:
```{r chunk5}
table(Default$default)
table(Default$default)/N*100 
```

And afterwards we split the data into two subsets - the learning set and the test set - containing 60% and 40% of units, respectively, using the function `sample()`

```{r chunk6}
totIndex=seq(from=1,to=N,by=1)
length(totIndex)
alpha=0.6
set.seed(100)
trainIndex=sample(totIndex,alpha*N, replace=F )
Default_L<-Default[trainIndex,] # Learning set
nL=nrow(Default_L)
Default_T<-Default[-trainIndex,] # Test set
nT=nrow(Default_T)
```

Now check the proportions of defaulters and non-defaulters in the two subsets and compare with the corresponding proportions in the complete dataset:

```{r chunk7}
table(Default$default)/N*100
table(Default_L$default)/nL*100
table(Default_T$default)/nT*100 
```

Consider another sample
```{r}
totIndex=seq(from=1,to=N,by=1)
length(totIndex)
alpha=0.6
set.seed(200)
trainIndex=sample(totIndex,alpha*N, replace=F )
Default_L<-Default[trainIndex,] # Learning set
nL=nrow(Default_L)
Default_T<-Default[-trainIndex,] # Test set
nT=nrow(Default_T)
table(Default$default)/N*100
table(Default_L$default)/nL*100
table(Default_T$default)/nT*100 
```
Even if these differences are small, using the `sample()` function we don't have any guarantee that the proportions of the values of the target variables will be maintained in either subset. 

We fix this drawback, using the `caret: Classification And REgression Training` package, see http://topepo.github.io/caret/data-splitting.html for details. In particular, we need the function *createDataPartition()*.

```{r chunk8}
library(caret)
set.seed(2000)
alpha_L= 0.6
trainIndex <- createDataPartition(Default$default, p = alpha_L, 
                                  list = FALSE, 
                                  times = 1)
Default_L<-Default[trainIndex,]  # Learning set
nL<-nrow(Default_L)
table(Default_L$default)
table(Default_L$default)/nL*100
```

The remaining units are obtained using  the `-trainIndex`  below selects  only the observations that are not in the learning  set.

```{r chunk9}
Default_VT<-Default[-trainIndex,]
nVT<-nrow(Default_VT)
table(Default_VT$default)
table(Default_VT$default)/nVT*100 
```

Afterwards we can split the remaining units into two further subsets,  the validation set and the test set having the same size (proportion=0.2):

```{r chunk10}
alpha_V=0.2 # the proportion of units in the validation set 
validIndex <- createDataPartition(Default_VT$default, p = alpha_V/(1-alpha_L), 
                                  list = FALSE, 
                                  times = 1)

Default_V<-Default_VT[validIndex,]     # Validation set
Default_T<-Default_VT[-validIndex,]    # Test set

nV<-dim(Default_V)[1]
nT<-dim(Default_T)[1]
table(Default_V$default)
table(Default_V$default)/nV*100
table(Default_T$default)
table(Default_T$default)/nT*100
```

And we can easily check that the proportions of defaulters and non-defaulters are the approximately the same in the three subsets (learning, validation and test set).

## 4.3. The Validation Set Approach

We explore the use of the validation set approach in order to estimate the test error rates that result from fitting various linear models on the `Auto` data set, consisting of data frame with 392 observations on the following 9 variables:

*mpg*:  miles per gallon </br>
*cylinders*: Number of cylinders between 4 and 8 </br>
*displacement*: Engine displacement (cu. inches) </br>
*horsepower*: Engine horsepower </br>
*weight*: Vehicle weight (lbs.) </br>
*acceleration*: Time to accelerate from 0 to 60 mph (sec.) </br>
*year*: Model year (modulo 100) </br>
*origin*: Origin of car (1. American, 2. European, 3. Japanese) </br>
*name*: Vehicle name </br>


Select a sample of size 196 from the dataset:

```{r chunk11}
library(ISLR2)
attach(Auto)
set.seed(1)
train <- sample(392, 196)
```

We then use the `subset` option in `lm()` to fit a linear regression using only the observations corresponding to the training set.

```{r chunk12}
lm.fit <- lm(mpg ~ horsepower, data = Auto, subset = train)
```

We now  use
 the `predict()` function to estimate the response for all $392$ observations,  and
 we  use
  the `mean()` function to calculate the MSE of the $196$ observations in the validation set. Note that the `-train` index below selects  only the observations that are not in the training set.

```{r chunk13}
attach(Auto)
mx1=mean((mpg - predict(lm.fit, Auto))[-train]^2)
mx1
```

Therefore, the estimated test MSE for the linear regression fit is `r round(mx1,3)`. We can use polynomial regression where the relationship between the predictors and the response is of kind
\[ y_i = \beta_0 + \beta_1 x_i + \beta_2 x_i^2 + \beta_3 x_i^3 + \cdots + \beta_d x_i^d + \varepsilon_i \]
see the textbook Sec. 7.1 for details.

In `R` polynomial regression is implemented through the  the `poly()` function and estimate the test error for the quadratic and cubic regressions.

```{r chunk14}
lm.fit2 <- lm(mpg ~ poly(horsepower, 2), data = Auto, 
    subset = train)
mx2=mean((mpg - predict(lm.fit2, Auto))[-train]^2)
mx2
lm.fit3 <- lm(mpg ~ poly(horsepower, 3), data = Auto, 
    subset = train)
mx3=mean((mpg - predict(lm.fit3, Auto))[-train]^2)
mx3
```

These error rates are `r round(mx2,3)` and `r round(mx3,3)`, respectively (quite smaller than than the provios value). 
If we choose a different training set instead, then we will obtain somewhat different errors on the validation set.


```{r chunk15}
set.seed(2)
train <- sample(392, 196)
lm.fit <- lm(mpg ~ horsepower, subset = train)
mx1b=mean((mpg - predict(lm.fit, Auto))[-train]^2)
mx1b
lm.fit2 <- lm(mpg ~ poly(horsepower, 2), data = Auto, 
    subset = train)
mx2b=mean((mpg - predict(lm.fit2, Auto))[-train]^2)
mx2b
lm.fit3 <- lm(mpg ~ poly(horsepower, 3), data = Auto, 
    subset = train)
mx3b=mean((mpg - predict(lm.fit3, Auto))[-train]^2)
mx3b
```

Using this split of the observations into a training set and a validation set,
we find that the validation set error rates for the models with linear, quadratic, and cubic terms are `r round(mx1b,3)`, `r round(mx2b,3)`, and `r round(mx3b,3)`, respectively.

**Remark.** These results are consistent with our previous findings: a model that predicts `mpg` using a quadratic function of `horsepower` performs better than a model that involves only a linear function of `horsepower`, and there is little evidence in favor of a model that uses a cubic function of `horsepower`.

