---
title: "StatLearning 2024/25"
output: html_document
---

# Lab activity 6: Linear Methods for Classification - Part II


## 6.1. Linear Discriminant Analysis (LDA)

Consider again the `Default` data in the `ISLR2` library

```{r chunk1}
library(ISLR2)
attach(Default)
N<-nrow(Default)
str(Default)
```

First, take a look at the data

```{r}
pairs(Default, col="red", pch=20, cex=0.5)
plot(income, balance, pch=20, cex=0.5, col = ifelse(student=="Yes",'blue','gray'))
plot(income, balance, pch=20, cex=0.5, col = ifelse(default=="Yes",'red','gray'))
plot(balance, default, pch=20, cex=0.5, col = ifelse(student=="Yes",'red','gray'))
```

Now consider Linear Discriminant Analysis that is implemented in the `MASS: Support Functions and Datasets  support Venables and Ripley book "Modern Applied Statistics with S" trough the function `lda()`. 

```{r chunk2}
library(MASS)
lda.fit=lda(default~balance+student) 
lda.fit
```
Once we get the model, we can compute the predicted value using the `predict()` function and afterwards the confusion matrix. Finally, we compute the accuracy and the misclassification error.

```{r chunk3}
lda.predict=predict(lda.fit,Default)
head(cbind(lda.predict$x,lda.predict$posterior, lda.predict$class))
pred.class=lda.predict$class
table(pred.class,default)   # confusion matrix

confMat.LDA<-addmargins(table(pred.class,default))
confMat.LDA
accuracy.LDA=(confMat.LDA[1,1]+confMat.LDA[2,2])/N*100 # accuracy
accuracy.LDA  
Err.LDA=100-accuracy.LDA # misclassification error
Err.LDA
```

## 6.2. Comparison between LDA and logistic regression

Now fit the logistic regression, using the same predictors: balance and student.

```{r chunk4}
glm.fit=glm(default~balance+student, family=binomial, data=Default)
summary(glm.fit, digits=3)
glm.predict=predict.glm(glm.fit,Default)
glm.probs=predict.glm(glm.fit,type="response")
glm.pred=rep("No",N)
glm.pred[glm.probs>.5]="Yes" # assign "Up" for units with P(Y="Up")>0.5
table(glm.pred,Default$default) # confusion matrix
confMat.LR<-addmargins(table(glm.pred,Default$default))
confMat.LR
accuracy.LR=(confMat.LR[1,1]+confMat.LR[2,2])/N*100 # accuracy
accuracy.LR
Err.LR=100-accuracy.LR # misclassification error
Err.LR
```

And finally compare the results
```{r chunk5}
accuracy.LDA
accuracy.LR
```

A more accurate model choice requires the analysis on a training set and on a test set. First, select  the training and test sets. As usual, we use the `createDataPartition()` function in `caret` library:

```{r chunk6}
library(caret)
library(caret)
table(Default$default)
table(Default$default)/N*100

set.seed(1500)
alpha=0.7
trainIndex <- createDataPartition(Default$default, p = alpha, 
                                  list = FALSE, 
                                  times = 1)
Default_L<-Default[trainIndex,]
nL<-dim(Default_L)[1]
nL
table(Default_L$default)
table(Default_L$default)/nL*100

Default_T<-Default[-trainIndex,]
nT<-dim(Default_T)[1]
nT
table(Default_T$default)
table(Default_T$default)/nT*100 

```

First, fit the model on the learning set

```{r chunk7}
lda.fit_L=lda(default~balance+student, Data=Default_L)  
lda.predict_L=predict(lda.fit_L,Default_L)
pred.class_L=lda.predict_L$class
confMat.LDA_L=addmargins(table(pred.class_L,Default_L$default))
confMat.LDA_L
accuracy.LDA_L=(confMat.LDA_L[1,1]+confMat.LDA_L[2,2])/nL*100
accuracy.LDA_L
Err.LDA_L=100-accuracy.LDA_L
Err.LDA_L
```

and then compute the fitted model on the test set and compare the accuracies on the learning and test sets:

```{r chunk8}
lda.predict_T=predict(lda.fit_L,Default_T)
pred.class_T=lda.predict_T$class
confMat.LDA_T=addmargins(table(pred.class_T,Default_T$default))
confMat.LDA_T
accuracy.LDA_T=(confMat.LDA_T[1,1]+confMat.LDA_T[2,2])/nT*100
accuracy.LDA_T
Err.LDA_T=100-accuracy.LDA_T
Err.LDA_T

accuracy.LDA_T
accuracy.LDA_L
```

Now consider the logistic regression model on the same datasets. First consider the learning set

```{r chunk9}
glm.fit_L=glm(default~balance+student, family=binomial, data=Default_L)
summary(glm.fit_L, digits=3)
glm.probs_L=predict.glm(glm.fit_L,type="response")
glm.pred_L=rep("No",nL)
glm.pred_L[glm.probs_L>.5]="Yes" # assign "Up" for units with P(Y="Up")>0.5
table(glm.pred_L,Default_L$default) # confusion matrix
confMat.LR_L<-addmargins(table(glm.pred_L,Default_L$default))
confMat.LR_L
accuracy.LR_L=(confMat.LR_L[1,1]+confMat.LR_L[2,2])/nL*100
accuracy.LR_L
Err.LR_L=100-accuracy.LR_L
Err.LR_L
```

and afterwards, compute the predicted values on the test set and the accuracy:

```{r chunk10}
glm.probs_T=predict.glm(glm.fit_L, newdata=Default_T, type="response") # note the option newdata
glm.pred_T=rep("No",nT)
glm.pred_T[glm.probs_T>.5]="Yes" # assign "Up" for units with P(Y="Up")>0.5
table(glm.pred_T,Default_T$default) # confusion matrix
confMat.LR_T<-addmargins(table(glm.pred_T,Default_T$default))
confMat.LR_T
accuracy.LR_T=(confMat.LR_T[1,1]+confMat.LR_T[2,2])/nT*100
accuracy.LR_T
Err.LR_T=100-accuracy.LR_T
Err.LR_T
```


## 6.3. Analysis of the ROC curve and AUC

There are many packages in `R` for the scoring analysis based on the ROC curve and the AUC. Here we consider the package `ROCR:  Visualizing the Performance of Scoring Classifiers`, see e.g.  https://www.r-bloggers.com/a-small-introduction-to-the-rocr-package/
for details. `ROCR` is easy to use, with only three commands and reasonable default values for all optional parameters. 

Every classifier evaluation using ROCR starts with creating a prediction object using the `prediction()` procedure. This function is used to transform the input data (which can be in vector, matrix, data frame, or list form) into a standardized format.

Now compute a prediction object based on the predicted values, i.e. on the variable *glm.probs_T* of the logistic regression model.

```{r chunk11}
library(ROCR)
contrasts(default)
pred.t.LR_T=prediction(glm.probs_T, Default_T$default) # 
```

The procedure `performance()` generate a set of performance measure, see the ROCR package documentation for details. The option `measure="tpr"` in   means *true positive rate* and the option `measure="fpr"` in `x.measure` means *false positive rate*. We compute 

```{r chunk12}
perf.t.LR_T=performance(pred.t.LR_T, measure = "tpr", x.measure = "fpr") 
```

Finally  we can plot the ROC curve:

```{r chunk13}
plot(perf.t.LR_T,colorize=TRUE,lwd=2) 
plot(perf.t.LR_T,colorize=TRUE,lwd=2, print.cutoffs.at=c(0.2,0.5,0.8)) 
abline(a=0,b=1, lty=2)
```

The AUC can be computed using the *performance* function considering the option `measure="auc"`:
```{r chunk14}
LR_T.auc=performance(pred.t.LR_T, measure = "auc") 
LR_T.auc@y.values[[1]]
```

Anagously, we plot the AUC curve and compute the AUC for the linear discriminant results.

```{r chunk15}
# LDA
pred.t.LDA_T=prediction(lda.predict_T$posterior[,2], Default_T$default)
perf.t.LDA_T=performance(pred.t.LDA_T, measure = "tpr", x.measure = "fpr")
plot(perf.t.LDA_T,colorize=TRUE,lwd=2, print.cutoffs.at=c(0.2,0.5,0.8))
abline(a=0,b=1, lty=2)
LDA_T.auc=performance(pred.t.LDA_T, measure = "auc") 
LDA_T.auc@y.values[[1]]
```

### Case study: wine data classification

These data are the results of a chemical analysis of wines grown in the same region in Italy but derived from three different cultivars in Italy. The analysis determined the quantities of the following 13 chemical concentrations found in each of the three types of wines:

1) Alcohol 
2) Malic acid 
3) Ash 
4) Alcalinity of ash 
5) Magnesium 
6) Total phenols 
7) Flavanoids 
8) Nonflavanoid phenols 
9) Proanthocyanins 
10) Color intensity 
11) Hue 
12) OD280/OD315 of diluted wines 
13) Proline 

```{r}
data(wine, package='rattle')
attach(wine)
str(wine)
head(wine)
N<-nrow(wine)
pairs(wine[2:14], col=factor(wine$Type), pch=20, cex=0.3)
```

Now consider linear discriminant analysis

```{r}
wine.lda <- lda(Type ~ ., data=wine)
wine.lda
```

This means that the first discriminant function is a linear combination of the 
variables: \[ −0.403∗Alcohol+0.165∗Malic+...+−0.003∗Proline. \] 
For convenience, the value for each discriminant function 
(eg. the first discriminant function) are scaled so that their mean value 
is zero and its variance is one.

The “proportion of trace” that is printed when you type “wine.lda” 
(the variable returned by the lda() function) is the percentage 
separation achieved by each discriminant function. 
For example, for the wine data we get the same 
values as just calculated (68.75% and 31.25%).

```{r}
wine.lda.values <- predict(wine.lda)
wine.lda.values$posterior
wine.lda.values$class
```

Scatterplots of the Discriminant Functions: we can obtain a scatterplot of the best two discriminant functions, with the data points labelled by cultivar.

```{r}
plot(wine.lda.values$x[,1],wine.lda.values$x[,2],  cex=0.8,pch=20, xlab="LDA axis 1", ylab="LDA axis 2") # make a scatterplot
text(wine.lda.values$x[,1],wine.lda.values$x[,2],Type,cex=0.7,pos=4,col="red") # add labels
plot(wine.lda.values$x[,1],wine.lda.values$x[,2],cex=0.8, col=Type, pch=20, xlab="LDA axis 1", ylab="LDA axis 2") 
plot(wine.lda.values$x[,1],wine.lda.values$x[,2],cex=0.8, col=c("blue","red","green")[Type], pch=20, xlab="LDA axis 1", ylab="LDA axis 2") 
```

From the scatterplot of the first two discriminant functions, 
we can see that the wines from the three cultivars are well separated 
in the scatterplot. The first discriminant function (x-axis) separates 
cultivars 1 and 3 very well, but doesn’t not perfectly separate 
cultivars 1 and 3, or cultivars 2 and 3.

The second discriminant function (y-axis) achieves a fairly good 
separation of cultivars 1 and 3, and cultivars 2 and 3, although 
it is not totally perfect.


To achieve a very good separation of the three cultivars, it would be best to use both 
the first and second discriminant functions together, since the first discriminant function can separate cultivars 1 and 3 very well, and the second discriminant function can separate  cultivars 1 and 2, and cultivars 2 and 3, reasonably well. 

```{r}
wine.lda.values$class
confMat.LDA<-confusionMatrix(Type,wine.lda.values$class)$table
confMat.LDA
accuracy.LDA<-sum(diag(confMat.LDA))/N*100
accuracy.LDA
```

#### Analysis on training and test set

To begin with, prepare the training and the test sets.
```{r}
set.seed(1500)
alpha=0.7
trainIndex <- createDataPartition(wine$Type, p = alpha, 
                                  list = FALSE, 
                                  times = 1)
wine_L<-wine[trainIndex,]
nL<-nrow(wine_L)
nL
table(wine_L$Type)
round(table(wine_L$Type)/nL*100,2)

wine_T<-wine[-trainIndex,]
nT<-nrow(wine_T)
nT
table(wine_T$Type)
round(table(wine_T$Type)/nT*100,2) 
```

Now analyze the training data.

```{r}
lda.fit_L=lda(Type ~ ., data=wine_L)  
lda.predict_L=predict(lda.fit_L,wine_L)
pred.class_L=lda.predict_L$class
confMat.LDA_L=table(wine_L$Type,pred.class_L)
confMat.LDA_L
accuracy.LDA_L<-sum(diag(confMat.LDA_L))/nL*100
accuracy.LDA_L
```

and then fit the model on the test data

```{r}
lda.predict_T=predict(lda.fit_L,wine_T)
pred.class_T=lda.predict_T$class
confMat.LDA_T=table(wine_T$Type,pred.class_T)
confMat.LDA_T
accuracy.LDA_T<-sum(diag(confMat.LDA_T))/nT*100
accuracy.LDA_T
accuracy.LDA_L
```

Finally compare the results on both the training and the test sets

```{r}
accuracy.LDA_T
accuracy.LDA_L
```


## 6.4. Quadratic Discriminant Analysis 

Consider the MBA admission data that provide admission data for applicants to graduate schools in business. The objective is to use the GPA and GMAT scores to predict the likelihood of admission (admit, notadmit, and borderline).

*GPA*: Grade Point Average, it is a score used to evaluate your success during the entirety of your degree programme </br>
*GMAT* Graduate Management Admission Test, it plays an important role in the MBA Admission process. </br>

```{r chunk16}
url <- 'http://www.biz.uiowa.edu/faculty/jledolter/DataMining/admission.csv'
admit <- read.csv(url)
str(admit)
head(admit)
N<-nrow(admit)
adm=data.frame(admit)
```

Lets start by doing the LDA, notice that in this case we have 3 classes:

```{r chunk17}
library(MASS)
lda.fit=lda(De~.,adm)
lda.fit
lda.pred=predict(lda.fit, admit)
lda.class=lda.pred$class
table(lda.class,admit$De)
accuracy.LDA<-sum(diag(table(lda.class,admit$De)))/N*100
accuracy.LDA
```

Now using LDA to predict a new student. The results can be interpreted as follows: the new student belongs to *admit* with probability 0.518, to *border* with probability 0.482 and to *notadmit* with negligible probability

```{r chunk18}
predict(lda.fit,newdata=data.frame(GPA=3.21,GMAT=497))
```

Now perform QDA

```{r chunk19}
qda.fit=qda(De~.,admit)
qda.fit
qda.pred=predict(qda.fit, admit)
qda.class=qda.pred$class
table(qda.class,admit$De)
accuracy.QDA<-sum(diag(table(qda.class,admit$De)))/N*100
accuracy.QDA
accuracy.LDA
```
and compare the results on the unit $(GPA=3.21,GMAT=497$). Here QDA provides a clearer prediction than LDA:
 
```{r chunk20}
predict(qda.fit,newdata=data.frame(GPA=3.21,GMAT=497))
predict(lda.fit,newdata=data.frame(GPA=3.21,GMAT=497))
```

Which model is best? In this case, the sample size is quite small. Therefore
in order to answer this question, we evaluate the LDA by randomly 
selecting 60 of 85 students, estimating the parameters on the training data, and 
classifying the remaining 25 students of the holdout sample. We repeat this 100 times.

```{r}
N<-nrow(admit)
nL=60
nT=N-nL
rep=100
```

Consider first LDA:
```{r}
set.seed(123456789)
delta.lda=dim(rep)
for (k in 1:rep) {
  train=sample(1:N,nL)
  ## LDA
  lda.fit=lda(De~.,admit[train,])
  predict(lda.fit,admit[-train,])$class
  table.lda=table(admit$De[-train],predict(lda.fit,admit[-train,])$class)
  delta.lda[k]=(nT-sum(diag(table.lda)))/nT
}
mean.delta.lda=mean(delta.lda) # mean classification error LDA
mean.delta.lda 
```

and afterwards model data through QDA:

```{r}
set.seed(123456789)
delta.qda=dim(rep)
for (k in 1:rep) {
  train=sample(1:N,nL)
  ## quadratic discriminant analysis
  qda.fit=qda(De~.,admit[train,])
  predict(qda.fit,admit[-train,])$class
  table.qda=table(admit$De[-train],predict(qda.fit,admit[-train,])$class)
  delta.qda[k]=(nT-sum(diag(table.qda)))/nT
}
mean.delta.qda=mean(delta.qda) # mean classification error LDA
mean.delta.qda 
```



